---
title: "Day 1: Community & Governance"
editor_options: 
  chunk_output_type: console
execute:
  echo: false
  warning: false
  message: false
---

```{r}
library(tidyverse)

agenda <- read_csv(here::here("data/tbl-openwashdata-conference-agenda.csv")) 

agenda_final <- agenda  |>
  select(-start_time, -end_time, -speaker_link, -affiliation_link) |>
  relocate(time) |>
  mutate(speaker = case_when(
    str_detect(speaker, "NA") == TRUE ~ " - ",
    TRUE ~ speaker
  )) |>
  mutate(affiliation = case_when(
    str_detect(affiliation, "NA") == TRUE ~ " - ",
    TRUE ~ affiliation
  ))
```

## Pre-requisites

- **Laptop**: Required for shared note-taking, especially in the afternoon sessions.
- **Google Doc access**: We will take notes in a shared Google Doc. You will receive an email to test if you can access it. If you did not receive an email, please contact the organizers.
- **Ideas for Hackathon**: Think about potential projects or topics you or somene else could  work on during the hackathon on Day 2.

## Schedule Overview

```{r}
agenda_final |> 
    filter(day == 1) |> 
    select(-day, -speaker) |> 
    rename_with(~stringr::str_to_title(.x), everything()) |> 
    knitr::kable() 
```

## Session Details

```{r}
agenda_sessions <- agenda_final |> 
  filter(day == 1) |> 
  select(-day, -speaker) |> 
  mutate(title = case_when(
    str_detect(title, "Warm-up activity") ~ "[Warm-up activity](warmup/)",
    str_detect(title, "openwashdata community") ~ "[openwashdata community - past and future](openwashdata/)",
    str_detect(title, "Governance Part 1") ~ "[Governance Part 1](governance/)",
    str_detect(title, "Governance Part 2") ~ "[Governance Part 2](governance/)",
    str_detect(title, "Hackathon Ideation") ~ "[Hackathon Ideation](ideation/)",
    str_detect(title, "Buffer") & str_detect(title, "Hackathon") ~ "[Hackathon Ideation](ideation/)",
    TRUE ~ title
  )) |> 
  rename_with(~stringr::str_to_title(.x), everything())

knitr::kable(agenda_sessions)
``` 



