---
title: "Day 1: Community & Governance"
editor_options: 
  chunk_output_type: console
execute:
  echo: false
  warning: false
  message: false
---

```{r}
library(tidyverse)

agenda <- read_csv(here::here("data/tbl-openwashdata-conference-agenda.csv")) 

agenda_final <- agenda  |>
  select(-start_time, -end_time, -speaker_link, -affiliation_link) |>
  relocate(time) |>
  mutate(speaker = case_when(
    str_detect(speaker, "NA") == TRUE ~ " - ",
    TRUE ~ speaker
  )) |>
  mutate(affiliation = case_when(
    str_detect(affiliation, "NA") == TRUE ~ " - ",
    TRUE ~ affiliation
  ))
```

## Pre-requisites

- **Laptop**: Required for shared note-taking, especially in the afternoon sessions.
- **Google Doc access**: We will take notes in a shared Google Doc. You will receive an email to test if you can access it. If you did not receive an email, please contact the organizers.
- **Ideas for Hackathon**: Think about potential projects or topics you or somene else could  work on during the hackathon on Day 2.

## Schedule Overview

```{r}
agenda_final |> 
    filter(day == 1) |> 
    select(-day) |> 
    rename_with(~stringr::str_to_title(.x), everything()) |> 
    knitr::kable() 
```

## Session Details

| Time | Title | Speaker | Affiliation |
|------|-------|---------|-------------|
| 09:00 - 09:30 | Arrival at Villa Hatt. Welcome coffee, bread rolls, croissants (Gipfeli). | - | - |
| 09:30 - 10:00 | [Warm-up activity](warmup/) | Lars | - |
| 10:00 - 10:45 | [openwashdata community - past and future](openwashdata/) | Lars | - |
| 10:45 - 11:00 | Break | - | - |
| 11:00 - 12:00 | Data stewardship at BASEflow Malawi | Emmanuel | - |
| 12:00 - 13:00 | Lunch break | - | - |
| 13:00 - 13:30 | Warm-up - Positive Gossip | - | - |
| 13:30 - 14:00 | [Governance Part 1](governance/) | Lars | - |
| 14:00 - 14:15 | Break | - | - |
| 14:15 - 15:00 | [Governance Part 2](governance/) | Lars | - |
| 15:00 - 15:30 | [Hackathon Ideation](ideation/) | - | - |
| 15:30 - 15:45 | Break with small cakes | - | - |
| 15:45 - 16:30 | [Lean Coffee](lean-coffee/) | - | - | 



