---
title: "Day 1: Community & Governance"
editor_options: 
  chunk_output_type: console
execute:
  echo: false
  warning: false
  message: false
---

```{r}
library(tidyverse)

agenda <- read_csv(here::here("data/tbl-openwashdata-conference-agenda.csv")) 

agenda_final <- agenda  |>
  select(-start_time, -end_time, -speaker_link, -affiliation_link) |>
  relocate(time) |>
  mutate(speaker = case_when(
    str_detect(speaker, "NA") == TRUE ~ " - ",
    TRUE ~ speaker
  )) |>
  mutate(affiliation = case_when(
    str_detect(affiliation, "NA") == TRUE ~ " - ",
    TRUE ~ affiliation
  ))
```

## Pre-requisites

- **Laptop**: Required for shared note-taking, especially in the afternoon sessions.
- **Google Doc access**: We will take notes in a shared Google Doc. You will receive an email to test if you can access it. If you did not receive an email, please contact the organizers.
- **Ideas for Hackathon**: Think about potential projects or topics you or somene else could  work on during the hackathon on Day 2.

## Schedule Overview

```{r}
agenda_final |> 
    filter(day == 1) |> 
    select(-day) |> 
    rename_with(~stringr::str_to_title(.x), everything()) |> 
    knitr::kable()
```

## Session Details

### Morning Sessions

The day begins with a warm-up session led by Lars, followed by a reflection on the openwashdata community's journey and future direction. Emmanuel will share insights on data stewardship implementation at BASEflow Malawi.

### Afternoon Sessions

After lunch, Liz will present on the intersection of openwashdata and Open Research Data from the Global Health Engineering perspective. 

The afternoon continues with:

- **Positive Gossip Exercise** - An energizing warm-up activity to reconnect after lunch
- **Governance Discussion** - Reflection on different governance style and how we might shape the future structure and decision-making processes of openwashdata
- **Hackathon Ideation** - Preparing a long list of ideas for the hackathon on Day 2
- **Lean Coffee** - An open, democratically-driven discussion on topics that matter most to the community



