---
title: "openwashdata conf 2025"
subtitle: "where do we head?"
author: 
  - name: "Lars Schöbitz"
    url: https://openwashdata.org/about/lars/
    affiliation: Global Health Engineering, ETH Zurich
    affiliation_url: https://ghe.ethz.ch/
date: 2025-07-15
date-format: long
format: 
  revealjs:
    logo: "../images/OWD-logo-20.png"
    theme: "../../style/slides.scss"
    transition: fade
    slide-number: true
    preview-links: auto
editor: source
editor_options: 
  chunk_output_type: console
---

```{r}

library(washopenresearch)
library(tidyverse)
library(ggthemes)
library(ggtext)
library(gt)

# set ggplot theme --------------------------------------------------------

ggplot2::theme_set(
    # Pick a starting theme
    ggplot2::theme_gray(base_size = 14) +
        # Add your favourite elements
        ggplot2::theme(
            axis.line = element_line(),
            panel.background = element_rect(fill = "white"),
            panel.grid.major = element_line("grey95", linewidth = 0.25),
            legend.key  = element_rect(fill = NA) 
        )
)

```


```{r}
agenda <- read_csv(here::here("data/tbl-openwashdata-conference-agenda.csv")) 

agenda_final <- agenda  |>
  select(-start_time, -end_time, -speaker_link, -affiliation_link) |>
  relocate(time) |>
  mutate(speaker = case_when(
    str_detect(speaker, "NA") == TRUE ~ " - ",
    TRUE ~ speaker
  )) |>
  mutate(affiliation = case_when(
    str_detect(affiliation, "NA") == TRUE ~ " - ",
    TRUE ~ affiliation
  ))
```


```{r}

# data preparation

washdev_das_type <- washdev |> 
    mutate(das_policy = case_when(
        published_year < 2020 ~ "pre-2020",
        TRUE ~ "2020 or later"
    )) |> 
    mutate(das_type = case_when(
        das_type == "in paper" ~ "available in paper",
        das_type == "on request" ~ "available on request",
        TRUE ~ das_type
    ))  |>     
    mutate(das_type = case_when(
        is.na(das_type) ~ "missing",
        TRUE ~ das_type
    )) 
```


```{r}

## summary for data availability statement (DAS) type and policy year

washdev_das_type_n <- washdev_das_type |> 
    count(das_policy, das_type) 


fig_das_type <- washdev_das_type_n |> 
    ggplot(aes(x = reorder(das_type, n), y = n, fill = das_policy)) +
    geom_col(position = position_dodge(), width = 0.6) +
    geom_text(aes(label = n), 
              vjust = 0.5, 
              hjust = -0.5,  
              size = 3,
              position = position_dodge(width = 0.5)
    ) +
    coord_flip() +
    annotate("text", 
             x = 3.77, 
             y = 150, 
             size = 3, 
             label = "after introducing policy\nfor data availability statement", 
             color = "gray20") +
    geom_curve(aes(x = 3.95, y = 142, xend = 3.95, yend = 70), 
               curvature = 0.5, 
               arrow = arrow(type = "closed", length = unit(0.1, "inches")),
               color = "gray20") +
    labs(
        title = "Data Availability Statement",
        subtitle = "Analysis of 924 articles published in Journal of Water, Sanitation and Hygiene for Development (2011 to 2023)",
        fill = "published year",
        y = "number of publications",
        x = "data availability statement") +
    scale_y_continuous(breaks = seq(0, 600, 100), limits = c(0,600)) +
    scale_fill_colorblind() +
    theme(panel.grid.major.y = element_blank(),
          plot.subtitle = element_text(size = 10))

# https://www.iwapublishing.com/news/iwa-publishing-2020-annual-review
```

```{r}
## summary for data availability statement (DAS) type and supp file type

washdev_supp_file_type_n <- washdev_das_type |> 
    filter(das_policy == "2020 or later") |> 
    select(paperid, das_type, supp_file_type) |> 
    unnest(supp_file_type) |> 
    mutate(supp_file_type = case_when(
        is.na(supp_file_type) ~ "missing",
        TRUE ~ supp_file_type
    )) |>
    count(das_type, supp_file_type) 

tbl_supp_type <- washdev_supp_file_type_n |> 
    group_by(supp_file_type) |> 
    summarise(n = sum(n)) |> 
    arrange(desc(n)) |> 
    mutate(perc = n / sum(n) * 100) 

```

# openwashdata conf 2025 {background-color="#5b195b"}

## Programme {.smaller .scrollable}

```{r}
agenda_final |> 
  filter(day == 1) |> 
  select(-day, -affiliation) |> 
  rename_with(~stringr::str_to_title(.x), everything()) |> 
  knitr::kable()
```

## {{< fa route >}} Session Roadmap (45 min)

::: incremental
- {{< fa chart-line >}} **The Opportunity** (~10 min) - Why open data matters in WASH
- {{< fa users >}} **openwashdata Journey** (~10 min) - Our story and framework
- {{< fa lightbulb >}} **Activity 1** (5 min) - Your strategies & tactics
- {{< fa rocket >}} **Future Plans** (~10 min) - WP2/WP3 & ds4owd-002
- {{< fa coins >}} **Funding & Collaboration** (~5 min)
- {{< fa users >}} **Activity 2** (5 min) - Building together
- {{< fa flag-checkered >}} **Wrap-up** (~2 min)
:::

::: {.notes}
This roadmap helps participants know what to expect and where we are in the journey
:::


# {{< fa chart-line >}} The Opportunity {background-color="#5b195b"}

::: {.notes}
⏱️ ~10 minutes for The Opportunity section
- We have a huge and missed opportunity in our sector
- Very little data is shared publicly and following best practices for reuse
:::

## Journal articles

```{r}
#| fig-width: 10
#| out-width: 100%
fig_das_type
```

```{r}
washdev_das_type_2020 <- washdev_das_type |> 
    filter(das_policy == "2020 or later") |> 
    count(das_type) |> 
    mutate(perc = n / sum(n) * 100) 

perc_missing <- washdev_das_type_2020 |> 
    filter(das_type == "missing") |> 
    pull(perc)

perc_in_paper <- washdev_das_type_2020 |> 
    filter(das_type == "available in paper") |> 
    pull(perc)
```


::: footer
Data: R package washopenresearch to be published at <https://github.com/openwashdata/washopenresearch>
:::

::: {.notes}
- The first missed opportunity are journal articles and data from researchers

- We looked at the Data Availability Statements in 924 articles published in the Journal of Water, Sanitation and Hygiene for Development from 2011 to 2023.

- You can see the data availability statements on the vertical axis and the number of publications on the horizontal axis

- Colors differentiate between papers published before 2020 and in 2020 or later, when a policy was introduced that requires authors to select one of the three data availability statements

- After that policy was introduced, we still found 15% of papers without a data availability statement, while 60% of articles stated that data was available in the paper, which could also be as supplementary material

:::

## Journal articles {.smaller}

#### Supplementary Material

:::: {.columns}

::: {.column width="60%"}

[Take-away]{.highlight-yellow}: Not a single file is in [machine-readable]{.highlight-yellow}, [non-proprietary]{.highlight-yellow} file type format that would qualify for following FAIR principles for data sharing [@wilkinson2016].

[Good practice]{.highlight-yellow}: CSV file (comma-separated values), including a data dictionary for all variables/columns in the data
:::

::: {.column width="40%"}
```{r}

tbl_supp_type |> 
    gt() |> 
    tab_header(title = "Supplementary Material",
               subtitle = "Articles published 2020 or later") |>
    tab_style(locations = cells_column_labels(), 
              style = cell_text(weight = "bold")) |>
    fmt_number(columns = c(perc), decimals = 1) |> 
    cols_label(supp_file_type = "file type", n = "n", perc = "%") |> 
    tab_footnote(
        footnote = md("One article can have multiple files."),
        locations = cells_column_labels(columns = n)
        )

```
:::

::::

::: footer
Data: R package washopenresearch to be published at <https://github.com/openwashdata/washopenresearch>
:::

::: {.notes}

- We then looked at the Supplementary Material of all articles published in 2020 or later and found that have half of the published articles still had no data published alongside the article

- But, the most insightful take-away is that not a single file was shared in a file type format that would qualify for following FAIR principles for data sharing.

- That is something we are hoping to change, where sharing data as CSV files would already go a long way.

:::

# {{< fa users >}} openwashdata community {background-color="#5b195b"} 

::: {.notes}
⏱️ START: 0:13 | DURATION: 3 minutes | END: 0:16
- To address this, we started the openwashdata community
- Quick overview of launch and support
:::

## openwashdata community

- Launched 10 March 2023
- Supported by four projects worth 340'000 CHF (50% in-kind contribution)
- Ends in July 2026
- So many outputs to write about (from 2026 to 2027)

## {{< fa eye >}} [Vision]{.highlight-yellow}

An active global community that applies FAIR principles [@wilkinson2016] to data generated in the greater water, sanitation, and hygiene sector.

## {{< fa bullseye >}} [Mission]{.highlight-yellow}

Empower WASH professionals to engage with tools and workflows for open data and code.

::: footer

From: [openwashdata.org/pages/gallery/vmost/](https://openwashdata.org/pages/gallery/vmost/)

:::

# {{< fa flag >}} VMOST as a method {background-color="#5b195b"}

## VMOST

- {{< fa eye >}} [V]{.highlight-yellow}ision
- {{< fa bullseye >}} [M]{.highlight-yellow}ission
- {{< fa list-check >}} [O]{.highlight-yellow}bjectives
- {{< fa chess >}} [S]{.highlight-yellow}trategy
- {{< fa tools >}} [T]{.highlight-yellow}actics

::: {.notes}
⏱️ START: 0:16 | DURATION: 7 minutes | END: 0:23
VMOST analysis is a tool used to evaluate if an overall strategy and supporting activities are in alignment. It can be used for current or future plans, and it breaks down a strategy and its core components into an easy-to-consume format
- Vision & Mission: 2 minutes
- Objectives: 2 minutes  
- Strategies & Tactics: 3 minutes
:::

## Objectives (Indicators) {.scrollable .smaller}

### By the end of March 2024 

::: incremental

1. Increase the number of datasets published on the website to [20 R data packages]{.highlight-yellow}.
1. Increase the number of datasets that are donated for publication to [50 datasets]{.highlight-yellow}.
1. Increase the number of people that have donated, cleaned, and published data independently with support of the openwashdata team to [5]{.highlight-yellow}.
1. Increase the number of unique visitors to the website to [10 visitors/day]{.highlight-yellow}.
1. Increase global coverage of visitors to the website to [50% of countries globally]{.highlight-yellow}.
1. Increase the number of data users who report having used data published through openwashdata community to [2 uses per dataset on average]{.highlight-yellow}.
1. Increase the number of subscribers to the openwashdata newsletter to[ 250 subscribers from 50 countries]{.highlight-yellow}.
1. Increase the number of participants in live coding events to [5 participants on average]{.highlight-yellow}. 

:::

## Strategies {.scrollable .smaller}

::: incremental

- Develop and maintain a data warehouse on the openwashdata website that provides an overview of published datasets.
- Develop a guide as a companion to workshops, live coding events, etc. that documents how to participate in the community and publish data following FAIR principles.
- Build a cohort of students, scientists, practitioners, and civil servants, that are comfortable using R statistical software for exploratory data analysis and Git version control and GitHub for communication and collaboration.
- Prepare all communication material for openwashdata using Quarto publishing framework^[https://quarto.org/] and R statistical software.
- Provide tools and resources to promote the use of open data in the WASH sector
- Publish workshops as open educational material.
- Introduce people to the concept of "open by default", as well as the use of open source software, the concept of open science, and benefits of open government (data).
- Build material always in mind with learner personas3 that were defined for the community.
- Communication material does not refer to openwashdata as a project, but as a community.
- Design a common corporate identity using defined color palettes, fonts, etc.
- Ensure that material developed for openwashdata follows best practices for accessibility.

:::

::: footer

Check: [openwashdata.org/pages/gallery/vmost/](https://openwashdata.org/pages/gallery/vmost/)

:::

## Tactics {.scrollable .smaller}

::: incremental

- Provide a 10-week online workshop for a selected group of participants to share tools and workflows that support publishing of open data following FAIR principles.
- Publish monthly blog posts on the openwashdata website about selected open datasets, community stories, workflows, insights into community management, use cases, etc.
- Publish monthly issues of the openwashdata newsletter.
- Host quarterly community meetups with invited speakers that share stories from their organisations related to data management, data analysis, open data, etc.
- Visualize and disseminate published open data using interactive dashboards, maps, articles, etc.

:::

::: footer

Check: [openwashdata.org/pages/gallery/vmost/](https://openwashdata.org/pages/gallery/vmost/)

:::

## {{< fa hand-paper >}} Get Ready to Share! {background-color="#e8f4f8"}

::: {style="font-size: 1.5em; text-align: center; margin-top: 2em;"}
In a moment, I'll pause to hear **your** ideas
:::

::: incremental
- {{< fa edit >}} Think about strategies & tactics for your context
- {{< fa globe >}} Consider your (local) challenges & opportunities  
- {{< fa users >}} What partnerships could accelerate progress?
:::

::: {.notes}
⏱️ This transition slide prepares participants for the upcoming activity
- Give them a moment to think before opening the Google Doc
- Emphasize we want practical, actionable ideas
:::

## {{< fa lightbulb >}} Activity: (5 minutes)

**In our shared Google Doc, please add:**

What other strategies or tactics could help build an open data culture in WASH?

Think about:

- 🎯 Your (local) context
- 🤝 Partnerships you could leverage
- 🛠️ Tools or resources you need

::: {.notes}
⏱️ START: 0:23 | DURATION: 5 minutes | END: 0:28
- Give clear instructions (1 min)
- Silent writing time (3 min)
- Quick verbal share of 1-2 ideas (1 min)
This builds on the VMOST framework they just learned. 
Watch for innovative ideas that could expand our approach.
Highlight connections to their suggestions when discussing WP3.
:::

# {{< fa rocket >}} Future {background-color="#5b195b"}

## {{< fa gavel >}} WP2: Governance (what today is all about)

- Activity 2.1: Develop a governance structure for a community organization and decision-making processes.
- Activity 2.2: Form a sounding board comprising community members to provide directional feedback.
- Activity 2.3: Create a long-term funding strategy for the openwashdata community.

[Open question: What is the medium-term future of openwashdata and what does it look like?]{.highlight-yellow}

::: {.notes}
⏱️ START: 0:28 | DURATION: 5 minutes | END: 0:33
- WP2 Governance: 2 minutes
- WP3 Community expansion: 3 minutes
- Reference back to participant ideas from Activity 1
::: 

## {{< fa users-rays >}} WP3: Community expansion (What I need your support for)

- {{< fa graduation-cap >}} Activity 3.1: Offer advanced data science training and workshops to community members.
- {{< fa handshake >}} Activity 3.2: Develop a mentorship program to support new members in adopting ORD practices.
- {{< fa calendar-days >}} Activity 3.3: Organize community events to foster networking and collaboration.

[Priority: Strong focus on WP3 for the remainder of the project.]{.highlight-yellow}

## data science for openwashdata 002 {.smaller}

All efforts into the next iteration of the course.

::: {.notes}
⏱️ START: 0:33 | DURATION: 5 minutes | END: 0:38
- Course overview and goals: 2 minutes
- Communication campaign: 1.5 minutes
- Course preparation: 1.5 minutes
:::

- free, live, online, 9 module programme ([goal: new AI module]{.highlight-yellow})
- 200 registrations for 2023 iteration ([goal 2025: 500]{.highlight-yellow})
- 100 show-ups ([goal 2025: 250]{.highlight-yellow})
- 20 participants completed capstone project ([goal: 100]{.highlight-yellow})
- 5 participants published data packages ([goal: 50]{.highlight-yellow})
- next iteration: from 11th September 2025, sign-up link: <https://ee-eu.kobotoolbox.org/x/7V3qeDYD>

::: learn-more
[ds4owd-002.github.io/website/](https://ds4owd-002.github.io/website/)

```{=html}
<iframe src="https://ds4owd-001.github.io/website/" width="100%" height="300" style="border:2px solid #123233;" title="Website of data science for openwashdata coursew"></iframe>
```
:::

## ds4owd-002 communication campaign (strategy from early June)

::: incremental

- 🌕 restart monthly newsletter editions (now) 
- 🌕 publish a blog post on ds4owd-001 (July) (Thanks, Adriana)
- 🌗 start publishing a LinkedIn post every Thursday (openwashdata thursday) 
- 🌑 host an information event (late August)
- 🌑 host a series of workshops for `washr` / `fairenough` (from October 2025 to March 2026)

:::

## ds4owd-002 course preparation

::: incremental

- 🌕 Platform for access to recordings through authentication (Zoom with registration for each session and detailed usage statistics)
- 🌗 Prepare quizzes for each module for participants to complete each module (with feedback on whether participants watched recording or joined live module)
- 🌗 Share information about course through personal channels, newsletters, etc.
- 🌑 Write templates / R function / Claude Code Slash commands for reviewing homework assignments, capstone projects, data packages, etc.
- 🌑 Establish mentorship programme

:::

# {{< fa sitemap >}} Organogram (Phase 3?) {background-color="#5b195b"}

## Starting Point

![](../../images/openwashdata-organogram-01.svg)

::: footer
Read full proposal for Phase 2 at: [openwashdata.org/pages/gallery/proposal-02/](https://openwashdata.org/pages/gallery/proposal-02/)
:::

::: {.notes}
⏱️ START: 0:38 | DURATION: 2 minutes | END: 0:40
- 30 seconds per slide (4 slides = 2 minutes)
- Emphasize evolution and growth
- Quick visual journey through phases
:::

## Mid-development

![](../../images/openwashdata-organogram-05.svg)

::: footer
Full evolution available at: [openwashdata.org/pages/gallery/proposal-02/](https://openwashdata.org/pages/gallery/proposal-02/)
:::

## Community Integration

![](../../images/openwashdata-organogram-09.svg)

::: footer
Full evolution available at: [openwashdata.org/pages/gallery/proposal-02/](https://openwashdata.org/pages/gallery/proposal-02/)
:::

## Future Vision

![](../../images/openwashdata-organogram-10.svg)

::: footer
Read full proposal for Phase 2 at: [openwashdata.org/pages/gallery/proposal-02/](https://openwashdata.org/pages/gallery/proposal-02/)
:::


# {{< fa coins >}} Funding opportunities {background-color="#5b195b"}

## On the list

- 2025-09-01: [ETH Open Research Data Program](https://www.lib4ri.ch/sites/default/files/media/documents/Creating-and-publishing-oer-on-rdm.pdf) - Creating and publishing open educational resources on research data management. CHF 10,000 – CHF 25,000. (10 months)
- 2025-09-30: [Swiss Leading House Africa](https://swisstph.submittable.com/submit) - Call for Research Partnership Grant. CHF 15,000 – CHF 40,000 (3 to 12 months).
- 2025-10-31: [Congressi Stefano Franscini](https://csf.ethz.ch/). 2nd open(wash)data / conference in July 2027. 
- TBD: [ETH Foundation](https://ethz-foundation.ch/en/)

::: {.notes}
⏱️ START: 0:40 | DURATION: 3 minutes | END: 0:43
- List funding opportunities: 2 minutes
- Transition slide: 30 seconds
- Set up for Activity 2: 30 seconds
:::

## {{< fa pencil >}} Activity 2 Coming Up {background-color="#e8f4f8"}

::: {style="font-size: 1.2em; text-align: center; margin-top: 1.5em;"}
Prepare to share your knowledge about:
:::

:::: {.columns}
::: {.column width="50%"}
### {{< fa coins >}} Funding
What opportunities exist in your region/sector?
:::

::: {.column width="50%"}
### {{< fa handshake >}} Collaboration  
Who would you partner with?
:::
::::

::: {.notes}
⏱️ Quick transition to prepare them for the second activity
- This helps them start thinking before we open the Google Doc
:::

## {{< fa users >}} Activity: Building Together (5 minutes)

**In our shared Google Doc, please share:**

1. What other funding opportunities exist for open data initiatives?

2. Who would you want to collaborate with on an open(wash)data project?

::: {.notes}
⏱️ START: 0:43 | DURATION: 5 minutes | END: 0:48
- Instructions: 1 minute
- Writing time: 3 minutes  
- Quick share: 1 minute
This helps identify potential funding sources we haven't considered.
Also helps with networking - participants can see shared interests.
Use this to transition to papers/impact discussion.
:::

# {{< fa file-lines >}} Papers from mid-2026 {background-color="#5b195b"}

## 12 months, 4 papers {.smaller}

::: {.notes}
⏱️ START: 0:48 | DURATION: 2 minutes | END: 0:50
- Callback slide: 30 seconds
- Papers overview: 1 minute
- Thank you & closing: 30 seconds
- Allow buffer time for questions if ahead of schedule
:::

::: incremental

1. [Setting the baseline:]{.highlight-yellow} FAIR / Open Data practices in the WASH sector
2. [Increasing competency:]{.highlight-yellow} Data from two iterations of data science for openwashdata course
3. [Streamlining workflows:]{.highlight-yellow} Development of an R package for FAIR data publication (`washr` / `fairenough`)
4. [Tracking impact:]{.highlight-yellow} Analytics from published data packages
5. What else?

:::

## Thanks `r emo::ji("sunflower")`

This project was supported by the  [Open Research Data Program of the ETH Board](https://ethrat.ch/en/eth-domain/open-research-data/).

The slides were created via revealjs and Quarto: [https://quarto.org/docs/presentations/revealjs/](https://quarto.org/docs/presentations/revealjs/)

You can [view source code of slides on GitHub](https://github.com/openwashdata-conf/2025/blob/main/day1/openwashdata.qmd)

Or you can [download slides in PDF format](https://openwashdata-conf.github.io/2025/day1/openwashdata.pdf)

This material is licensed under [Creative Commons Attribution Share Alike 4.0 International](https://creativecommons.org/licenses/by-sa/4.0/).
