---
title: "Day 2: Claude Code Hackathon"
editor_options: 
  chunk_output_type: console
execute:
  echo: false
  warning: false
  message: false
---

```{r}
library(tidyverse)

agenda <- read_csv(here::here("data/tbl-openwashdata-conference-agenda.csv")) 

agenda_final <- agenda  |>
  select(-start_time, -end_time, -speaker_link, -affiliation_link) |>
  relocate(time) |>
  mutate(speaker = case_when(
    str_detect(speaker, "NA") == TRUE ~ " - ",
    TRUE ~ speaker
  )) |>
  mutate(affiliation = case_when(
    str_detect(affiliation, "NA") == TRUE ~ " - ",
    TRUE ~ affiliation
  ))
```

::: {.callout-important}
## Prerequisites Required
Please ensure you've completed all [prerequisite setup](../prerequisites/index.qmd) before arriving.
:::

## Session Details

```{r}
agenda_sessions <- agenda_final |> 
  filter(day == 2) |> 
  select(-day, -speaker, -affiliation) |> 
  mutate(title = case_when(
    str_detect(title, "Claude Code") ~ "[Why Claude Code - AI-Assisted Creation for GHE](slides/)",
    TRUE ~ title
  )) |> 
  rename_with(~stringr::str_to_title(.x), everything())

knitr::kable(agenda_sessions)
```


## {{< fa presentation >}} Slides

[View slides in full screen](slides.qmd) | [Download slides as PDF](https://github.com/openwashdata-conf/2025/raw/main/day2/why-claude-code-slides.pdf)

```{=html}
<iframe class="slide-deck" src="slides.html" height="420" width="747" style="border: 1px solid #123233;"></iframe>
```

## Resources

### Hackathon Handouts
- [Claude Code Workflow Cheatsheet](workflow-cheatsheet.pdf) - One-page reference guide for the complete workflow
- [Claude Code Tips Cheatsheet](tips-cheatsheet.pdf) - Quick reference for tips, tools, and best practices

### Documentation & Guides
- [Claude Code Overview](https://docs.anthropic.com/en/docs/claude-code/overview) - Official documentation
- [Greg Wilson's Blog: Tooling for Snailz](https://third-bit.com/2025/03/29/tooling-for-snailz/) - Real-world experience with Claude Code
- [PhD-level intelligence, or the graduate student's lament](https://clauswilke.substack.com/p/phd-level-intelligence-or-the-graduate) - Critical perspective on AI claims
- [Claude Directory](https://claude.ai/directory) - Discover and connect tools to Claude

### Critical Perspectives on AI Development
- [My AI Skeptic Friends Are All Nuts](https://fly.io/blog/youre-all-nuts/) - On the disconnect between AI skeptics and practitioners
- [Reddit Discussion: My AI Skeptic Friends Are All Nuts](https://www.reddit.com/r/artificial/comments/1l2138b/my_ai_skeptic_friends_are_all_nuts/) - Community discussion on AI expectations and context loss

### Tools & Integrations
- [ccusage](https://github.com/ryoppippi/ccusage) - Track your Claude Code token usage
- [Awesome MCP servers](https://github.com/punkpeye/awesome-mcp-servers) - Community-curated list of Model Context Protocol integrations
- [Lars's favorite MCPs](https://github.com/stars/larnsce/lists/mcps) - Curated MCP recommendations

### Video Tutorials
- [Getting Started with Claude Code](https://www.youtube.com/watch?v=6eBSHbLKuN0) - Official Anthropic tutorial
- [History and Concepts of Claude Code](https://www.youtube.com/watch?v=Lue8K2jqfKk) - Understanding the why behind Claude Code
- [Advanced Claude Code Workflows](https://www.youtube.com/watch?v=TiNpzxoBPz0&list=LL&index=2&t=4s) - Power user tips and tricks
- [Building Projects with Claude Code](https://www.youtube.com/watch?v=aHTXccrfXC8&t=1223s) - Real-world project walkthrough
